Temporal ordering and registration of images in studies of developmental dynamics

### 2014-08-19 image_analysis_paper2/pnas_draft.tex
added more for paper edits

%\graphicspath{ {paper_figures2/} }

@@

\significancetext{Cross-sectional imaging studies are commonly used to study developmental dynamics in multiple organisms. 
%
In these studies, a representative developmental trajectory is reconstructed from snapshots of different embryos, each of which has been arrested at a different point of its development. 
%
One of the first steps in this reconstruction involves registration and temporal ordering of images. 
%
Here we show that both of these tasks can be automated and combined in a single step using recent advances in data mining. 
%
To illustrate this approach, we temporally order and register several data sets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.}

@@

@@

%\begin{figure}[t]
%\includegraphics{fig1}
%\caption{Caricature illustrating the tasks of image registration and temporal ordering. {\xfigtextfontit (A)} Images of ``samples'', each in a different orientation and a different stage of development. {\xfigtextfontit (B)} Registered and ordered samples. For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
%%\label{fig:fish}
%\customlabel{fig:fish}{1}
%\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
%\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
%\end{figure}

@@

%\begin{figure}[t]
%\includegraphics{fig2}
%\caption{Schematic illustrating angular synchronization and diffusion maps. {\xfigtextfontit (A)} Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. {\xfigtextfontit (B)} The vectors from {\xfigtextfontit A}, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in {\xfigtextfontit A}: the differences between a pair of angles in {\xfigtextfontit B} is the same as the pairwise angle in {\xfigtextfontit A}. {\xfigtextfontit (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit \SI}), so that close data points are connected by darker (``stronger'') edges. {\xfigtextfontit (D)} The data in {\xfigtextfontit C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
%%\label{fig:schematics}
%\customlabel{fig:schematics}{2}
%\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
%\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
%\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
%\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
%\end{figure}

%\begin{figure*}[t]
%\includegraphics{fig3}
%\caption{{\xfigtextfontit Drosophila} embryos during cellularization. {\xfigtextfontit (A)} Images of embryos collected during cellularization. Each image is of a different embryo arrested at a different developmental time, and in a different rotational and translational orientation. {\xfigtextfontit (B)} Images from {\xfigtextfontit A}, now registered and ordered using vector diffusion maps. The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
%\customlabel{fig:data1}{3}
%\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
%\customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
%\customlabel{subfig:averaged_data1}{\ref{fig:data1}{\it C}}
%\end{figure*}

\subsection{Images of embryos during cellularization}

@@

%\begin{figure*}[t]
%\includegraphics{fig4}
%\caption{{\xfigtextfontit Drosophila} embryos during gastrulation. {\xfigtextfontit (A)} Images of gastrulating embryos. Each image is of a different embryo arrested at a different developmental time, and in a different translation and rotational orientation. {\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps. The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right. {\xfigtextfontit (C)} A representative ``developmental path'' from local averaging of the images. Each image is a (Gaussian-weighted) average of the images in {\xfigtextfontit B}. }
%\customlabel{fig:data2}{4}
%\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
%\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
%\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
%\end{figure*}

@@

\subsection{Identifying distinct developmental trajectories}

When time is the main source of variability, we have shown that vector diffusion maps can recover developmental progression from cross-sectional data.
%
However, vector diffusion maps can also provide an organization of data which contains multiple sources of variability.
%
To illustrate this point, we collected a data set consisting of $42$ embryos undergoing cellularization during the third hour of development (see \fig~\ref{fig:raw_data3}).
%
Half of the embryos are wild type, and half of the embryos are mutants where dpERK is expressed in one broad peak at the ventral side \cite{snail}.
%
Again, each image is an optical cross-section of an embryo fixed at a different developmental time, and the embryos have been stained for nuclei (gray), Dl (green), and dpERK (red).
%
We expect two sources of variability in this data set: differences due to developmental time, and differences between mutant and wild type dynamics.

We computed the vector diffusion maps embedding for our data, and the eigenvalue distribution (see \fig~\ref{fig:eigenvalues}) indicates that two coordinates may be required for a meaningful description of the data. 
%
\fig~\ref{subfig:projection_data3} shows the data projected onto the first two vector diffusion maps coordinates, where each point in this projection corresponds to one image in our data set.
%
Two distinct branches appear to emerge in this two-dimensional embedding space.

The inset of \fig~\ref{subfig:projection_data3} shows the two-dimensional projection, with the data points now colored to indicate whether they correspond to wild type or mutant embryos.
%
The two branches in the embedding help distinguish the two genotypes.
%
Selected images along the wild type and mutant trajectories are shown in \fig~\ref{subfig:average_wt_data3} and \fig~\ref{subfig:average_mut_data3}, respectively.
%
The portion of the embedding where the two genotypes are intermixed (lower-left region of the plot) contains embryos which are early in their developmental trajectory, where there is little dpERK signaling and the two groups of embryos are indistinguishable. 
%
When the two paths diverge, the wild type data develops two distinct peaks, 
%
%\begin{figure}[h]
%\includegraphics{fig5}
%\caption{Wild type and mutant {\xfigtextfontit Drosophila} embryos.  {\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~\ref{fig:raw_data3}), projected onto the first two vector diffusion maps coordinates. The directions of the developmental trajectories are indicated by arrows. (Inset) The projected data, now colored by genotype. The two branches separate mutant and wild type samples. {\xfigtextfontit (B)} dpERK signal at select points along the wild type trajectory in {\xfigtextfontit A} (denoted by circles in the inset). The two dpERK peaks which arise during development are indicated by arrows. {\xfigtextfontit (C)} dpERK signal at select points along the mutant trajectory in {\xfigtextfontit A} (denoted by circles in the inset). The single dpERK peak is indicated by an arrow. }
%\customlabel{fig:data3}{5}
%\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
%\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it B}}
%\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
%\end{figure}
%
\noindent whereas the mutant data develops a single peak.
%
Therefore, vector diffusion maps can also help distinguish different expression patterns and developmental trajectories.

@@

We presented a unified approach to temporal ordering and registration of images in cross-sectional studies of developmental dynamics. 
%
To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other. 
%
In particular, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \cite{yuan2014automated, surkova2008characterization}.  
%
In parallel, ordering of large-scale cross-sectional data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different chemical species \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}. 
%
Automated temporal ordering of such data sets was accomplished by first projecting the data onto a low-dimensional subspace spanned by the leading principal or independent components, and then solving a traveling salesman problem or constructing a minimum spanning tree on the projected data to order multiple snapshots. 
%
The approach presented here is different because it addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace; vector diffusion maps can also register images when global registration of the entire data set is topologically impossible \cite{zhao2014rotationally}.
%
The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information.

@@

\begin{figure}[h]
\includegraphics{fig5}
\caption{Wild type and mutant {\xfigtextfontit Drosophila} embryos.  {\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~\ref{fig:raw_data3}), projected onto the first two vector diffusion maps coordinates. The directions of the developmental trajectories are indicated by arrows. (Inset) The projected data, now colored by genotype. The two branches separate mutant and wild type samples. {\xfigtextfontit (B)} dpERK signal at select points along the wild type trajectory in {\xfigtextfontit A} (denoted by circles in the inset). The two dpERK peaks which arise during development are indicated by arrows. {\xfigtextfontit (C)} dpERK signal at select points along the mutant trajectory in {\xfigtextfontit A} (denoted by circles in the inset). The single dpERK peak is indicated by an arrow. }
\customlabel{fig:data3}{5}
\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it B}}
\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
\end{figure}

>>>>

\graphicspath{ {paper_figures2/} }

@@

%\significancetext{Cross-sectional imaging studies are commonly used to study developmental dynamics in multiple organisms. 
%%
%In these studies, a representative developmental trajectory is reconstructed from snapshots of different embryos, each of which has been arrested at a different point of its development. 
%%
%One of the first steps in this reconstruction involves registration and temporal ordering of images. 
%%
%Here we show that both of these tasks can be automated and combined in a single step using recent advances in data mining. 
%%
%To illustrate this approach, we temporally order and register several data sets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.}

@@

\section{Significance}
Cross-sectional imaging studies are commonly used to study developmental dynamics in multiple organisms. 
%
In these studies, a representative developmental trajectory is reconstructed from snapshots of different embryos, each of which has been arrested at a different point of its development. 
%
One of the first steps in this reconstruction involves registration and temporal ordering of images. 
%
Here we show that both of these tasks can be automated and combined in a single step using recent advances in data mining. 
%
To illustrate this approach, we temporally order and register several data sets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.
\vspace{0.5cm}

@@

@@

\subsection{Ordering Images From a Movie}

@@

@@

@@

We presented a unified approach to temporal ordering and registration of images in cross-sectional studies of developmental dynamics. 
%
To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other. 
%
In particular, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \cite{yuan2014automated, surkova2008characterization}.  
%
In parallel, ordering of large-scale cross-sectional data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different chemical species \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}. 
%
Automated temporal ordering of such data sets was accomplished by first projecting the data onto a low-dimensional subspace spanned by the leading principal or independent components, and then solving a traveling salesman problem or constructing a minimum spanning tree on the projected data to order multiple snapshots. 
%
The approach presented here is different because it addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace; vector diffusion maps can also order images when global registration of the entire data set is topologically impossible \cite{zhao2014rotationally}.
%
The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information.

@@

%\begin{figure}[t]
%\includegraphics{fig1}
%\caption{Caricature illustrating the tasks of image registration and temporal ordering. {\xfigtextfontit (A)} Images of ``samples'', each in a different orientation and a different stage of development. {\xfigtextfontit (B)} Registered and ordered samples. For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
%\label{fig:fish}
%\customlabel{fig:fish}{1}
%\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
%\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
%\end{figure}
%
%\begin{figure}[t]
%\includegraphics{fig2}
%\caption{Schematic illustrating angular synchronization and diffusion maps. {\xfigtextfontit (A)} Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. {\xfigtextfontit (B)} The vectors from {\xfigtextfontit A}, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in {\xfigtextfontit A}: the differences between a pair of angles in {\xfigtextfontit B} is the same as the pairwise angle in {\xfigtextfontit A}. {\xfigtextfontit (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit \SI}), so that close data points are connected by darker (``stronger'') edges. {\xfigtextfontit (D)} The data in {\xfigtextfontit C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
%\label{fig:schematics}
%\customlabel{fig:schematics}{2}
%\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
%\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
%\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
%\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
%\end{figure}
%
%\begin{figure*}[t]
%\includegraphics{fig3}
%\caption{{\xfigtextfontit Drosophila} embryos during cellularization. {\xfigtextfontit (A)} Images of embryos collected during cellularization. Each image is of a different embryo arrested at a different developmental time, and in a different rotational and translational orientation. {\xfigtextfontit (B)} Images from {\xfigtextfontit A}, now registered and ordered using vector diffusion maps. The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
%\customlabel{fig:data1}{3}
%\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
%\customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
%\customlabel{subfig:averaged_data1}{\ref{fig:data1}{\it C}}
%\end{figure*}
%
%\begin{figure*}[t]
%\includegraphics{fig4}
%\caption{{\xfigtextfontit Drosophila} embryos during gastrulation. {\xfigtextfontit (A)} Images of gastrulating embryos. Each image is of a different embryo arrested at a different developmental time, and in a different translation and rotational orientation. {\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps. The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right. {\xfigtextfontit (C)} A representative ``developmental path'' from local averaging of the images. Each image is a (Gaussian-weighted) average of the images in {\xfigtextfontit B}. }
%\customlabel{fig:data2}{4}
%\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
%\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
%\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
%\end{figure*}
%
%\begin{figure}[h]
%\includegraphics{fig5}
%\caption{Wild type and mutant {\xfigtextfontit Drosophila} embryos.  {\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~\ref{fig:raw_data3}), projected onto the first two vector diffusion maps coordinates. The directions of the developmental trajectories are indicated by arrows. (Inset) The projected data, now colored by genotype. The two branches separate mutant and wild type samples. {\xfigtextfontit (B)} dpERK signal at select points along the wild type trajectory in {\xfigtextfontit A} (denoted by circles in the inset). The two dpERK peaks which arise during development are indicated by arrows. {\xfigtextfontit (C)} dpERK signal at select points along the mutant trajectory in {\xfigtextfontit A} (denoted by circles in the inset). The single dpERK peak is indicated by an arrow. }
%\customlabel{fig:data3}{5}
%\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
%\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it B}}
%\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
%\end{figure}
